Building the Million Transcriptome Ocular Meta-Atlas

### 2021-01-20 manuscript_00.Rmd
updates

```{r Setup, message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(pool)
library(RSQLite)
library(formattable)
library(ggalluvial)
library(scattermore)
library(grid)
library(ComplexHeatmap)
library(viridis)
library(Matrix)
library(matrixStats)
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

# Introduction
## Explosion of single cell transcriptomic atlases
The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields. As of January 2021, over 40 million cells have been sequenced across over 1,200 studies and the average size of each study past 2020 is over 100,000 cells [@svenssonCuratedDatabaseReveals2020]. The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015] and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian organisms [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].

@@

```{r fig1, fig.width=13, fig.height=6, fig.cap=fig1_cap, echo=FALSE, message = FALSE}
fig1_cap <- fig_cap(name='fig1', caption =  'a. Counts of (study) accession, published papers, and batches for each scRNA technlogy, split by organism. b. Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods (SnakePOP) to produce batch corrected latent dimensions (LatentDims), then downstream analysis outputs like clustering, differential gene testing (DiffTesting), and 2D UMAP visualization. c. Cell type counts extracted from published studies for the more common retina cell types, split by species. Count of study accessions for each species overlaid on bar plot.')

source('figs_and_tables/fig1.R')
plot_grid(plot_grid(a, c, align = 'v',  axis = 'lr', ncol = 1, labels = c('','c')), b,  nrow = 1, rel_widths = c(1.1,1), labels = c('a','b'))
```

@@

## Running `r methods %>% length()` tools in a Snakefile-based system

@@

```{r figBenchmark, fig.width=14, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE}
figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a. Example of a method (combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type). b. no batch correction cleanly separates cell types but does not mix batches (colored by study). c. sumZScale (higher is better) for each method across a variety of data normalizations. All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions d. Optimal parameters for scVI methods across 1000 to 10000 HVGs and 4 to 100 latent dimensions.')

@@

```{r figXGboost_umap, fig.width=18, fig.height=13, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE}
figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  'filler text hello world')

@@

input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 0
input[['pt_size_meta']] <- 0
#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['label_toggle']] <- 2
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- NULL
ctp <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)

input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 1
input[['pt_size_meta']] <- 1
input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- c('AC/HC Precursors','Amacrine Cells','Astrocytes','Bipolar Cells','Cones','Early RPCs','Endothelial','Horizontal Cells','Late RPCs','Microglia','Muller Glia','Neurogenic Cells','Pericytes','PR Precursors','Red Blood Cells','Retinal Ganglion Cells','Rod Bipolar Cells','Rods','RPCs','RPE','Schwann','Tabula Muris')
ctp2 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                    celltype_predict_labels,
                                    celltype_labels,
                                    tabulamuris_predict_labels,
                                    cluster_labels,
                                    cat_to_color_df
)

input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 0
input[['pt_size_meta']] <- 0
#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- NULL
org <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)

@@

After ML projection of cell type labels from the original `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` labels onto a total of `r meta_filter %>% filter(!is.na(CellType_predict) | !is.na(TabulaMurisCellType_predict)) %>% nrow()` cells we have substantially improved the number of studies per cell type. For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labelled Muller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labelling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`). Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean()` after transferring the cell type labels.

As another check on the quality of the cell type assignments, we ran pairwise differential expression tests between the predicted cell types (see methods for further details). We show the five most consistently differentially expressed genes for each of the major retina cell types (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a). As a simple metric to identify known and unknown genes relating to the cell type specific expression we search PubMed for the number of publications with two searches per gene. We expect most of the genes identified to be known in the literature. The first search is the more precise "gene AND cell type" (e.g. "PDE6H AND Cones") and the second search is the more inclusive "gene AND retina" (e.g. "PDE6G AND Retina"). Of the 55 genes in (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`a), 40 had one or more citations in the gene - cell type search (table X) and 51 had one or more citation in the inclusive search. The 55 genes had a mean of `r top_markers %>% group_by(cluster)  %>% slice_max(mean_auc, n =5) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search). In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round`.

@@

The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space. The major remaining retina cell types, by proportion in the mammalian eye are the muller glia, which are a glial cell type which help support the neurons of the retina. Next we have the neural cell types which transmit and help interpret the signals from the photoreceptors before they leave the retina via the optic stalk: the amacrine cells, retinal ganglia, horizontal, and bipolar cells. All of these cells are in well defined spaces in the UMAP. Finally we see even across species that the major cell types still group near each other (`r supFig_cap(name = 'figXGboost_umap', display = 'cite')`d). The macaque retina cells are not present in the precursor/neurogenic center of the UMAP as expected because only fully developed tissues were sampled in these data.

@@

Many retina "atlases" have been published to date. We argue that we have created the first atlas that is generally useful because 1. our dataset/atlas is several times larger than any other published set, 2. our data is available via download in several forms at https://github.com/davemcg/scEiaD, and crucially 3. we provide a Google colab/Jupyter notebook which exactly lays out how to use scVI's to project (or query) outside data onto our scEiaD with minimum compute resources. Finally we also provide our xgboost model so outside groups can label their own datasets with our robust retina cell type labelling system.

# Supplemental
```{r supFig1, fig.width=4, fig.height=6, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE}
supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination. The right bar specifies the number of each cells in each post QC category. In silico doublets were identified with scrublet and DoubletDetector.')

Words

Spacing

```{r supFig_benchmark, fig.width=12, fig.height=8, out.width=300, fig.cap=supFig_benchmark_cap, echo=FALSE, message = FALSE}
supFig_benchmark_cap <- supFig_cap(name='supFig_benchmark', caption =  'Performance of the various batch correction tools across various benchmarking metrics. For the LISI and Silhouette plots in A, B higher means better batch mixing and further to the right means better cluster purity. For the ARI and NMI metrics (which reflects how well cluster matches with cell type) in C, D, higher means a better score.')

@@

Words

Spacing

```{r supTab1_CTcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab1_CTcounts_cap, echo=FALSE, message = FALSE}
supTab1_CTcounts_cap <- supTab_cap(name='supTab1_CTcounts_cap', caption =  'Counts for cell type labels. Published are the author created labels from the published datasets. Transferred area the cell labels that were transferred by a machine learning model onto the entire scEiaD dataset.')

source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
fontsize(ctTable, size = 8, part = 'all')
```

Words

Spacing

```{r supTab_SAcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab_SAcounts_cap, echo=FALSE, message = FALSE}
supTab_SAcounts_cap <- supTab_cap(name='supTab_SAcounts_cap', caption =  'Counts for number of studies with cell types labels before and after cell type label transfer')

source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
fontsize(saTable, size = 8, part = 'all')
```

Words

Spacing

```{r supFig_PR, fig.width=12, fig.height=10,  fig.cap=supFig_PR_cap, echo=FALSE, message = FALSE}
supFig_PR_cap <- supFig_cap(name='supFig_PR', caption =  'Precision recall curves for our xgboost cell type predictor model across each cell type predicted')

source('figs_and_tables/sup_fig__xgboost_PR.R')
pr_split_curves
```

Words

Spacing

```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, echo=FALSE, message = FALSE}
supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'Area under the precision recall curve (AUC) for each cell type, split by study')

#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 3), size = 8, part = 'all')
```

Words

Spacing

```{r supTab_CTmislabels, fig.width=2, fig.cap=supTab_CTmislabels_cap, echo=FALSE, message = FALSE}

supTab_CTmislabels_cap <- supTab_cap(name='supTab_CTmislabels', caption =  'TEXT2')

#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
fontsize(predictions %>%  
           mutate(max_pred_prob = rowMaxs(.[,-(31:ncol(predictions))] %>% as.matrix )) %>% 
           select(-cell_type_id) %>% 
           rename(PredCellType = CellType) %>% 
           inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
           filter(!is.na(TrueCellType)) %>% 
           mutate(pred_correct  = ifelse(PredCellType == TrueCellType, 'Correct', 'incorrect')) %>% 
           group_by(PredCellType, TrueCellType) %>% 
           summarise(Count = n()) %>% 
           mutate(Ratio  = round(Count / sum(Count),2 )) %>% 
           filter(Ratio > 0.05) %>% 
           filter(!grepl('Doubl|Periocu', TrueCellType)) %>% 
           flextable() %>% 
           autofit(), 
         size = 8, 
         part = 'all')
```

Words

Spacing

```{r supFig_platformInt, fig.width=12, fig.height=10,  fig.cap=supFig_platformInt_cap, echo=FALSE, message = FALSE}
supFig_platformInt_cap <- supFig_cap(name='supFig_platformInt', caption =  '')
input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 5
input[['pt_size_meta']] <- 5
input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
#input[['meta_filter_cat']] <- 'CellType_predict'
#input[['meta_filter_on']] <- 'Bipolar Cells'
ctp3 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                    celltype_predict_labels,
                                    celltype_labels,
                                    tabulamuris_predict_labels,
                                    cluster_labels,
                                    cat_to_color_df
)

@@

```{r projection_fig, fig.width=12, fig.height=10,  fig.cap=projection_fig_cap, echo=FALSE, message = FALSE}
projection_fig_cap <- supFig_cap(name='projection_fig', caption =  '')
input[['pt_size_back']] <- 0
input[['pt_size_meta']] <- 6
input[['meta_column']] <- 'study_accession'
#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- c('RPE')
rpe <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)
#

@@

>>>>

```{r Setup, message=FALSE, warning=FALSE, include=FALSE}
#knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
library(tidyverse)
library(citr)
library(cowplot)
library(ggrepel)
library(colorspace)
library(flextable)
library(captioner)
library(pool)
library(RSQLite)
library(formattable)
library(ggalluvial)
library(scattermore)
library(grid)
library(ComplexHeatmap)
library(viridis)
library(Matrix)
library(matrixStats)
# setup caption-ing
fig_cap <- captioner("Figure")
supFig_cap <- captioner("Supplemental Figure")
#tab_cap <- captioner("Table")
supTab_cap <- captioner("Supplemental Table")

@@

# Introduction

## The retina, single cell gene analysis, and a recent explosion

The retina contains a multitude of cell types that, in total, are responsible for turning light information into signal for the brain to interpret as vision. Very briefly, the photoreceptors (rods and cones) are responsible for capturing the photons. The retinal pigmented epithelium (RPE) behind the photoreceptors physically support the rods and cones by processing byproducts of the visual cycle. Müller glia serve as support cells for the neurons. The retinal bipolar cells transmit the electrical signal from the photoreceptors to the retinal ganglion cells. Horizontal and amacrine regulate and help interpret signals from the photoreceptors. The signal is relayed via the retinal ganglion projections through the optic stalk to the brain (for review see [@maslandNeuronalOrganizationRetina2012]). Since 2000 many groups have investigated gene expression in small numbers of individual cells of the retina [@anniesIsoformPatternAChR2002a; @hagstromConePigmentGene2000; @trimarchiMolecularHeterogeneityDeveloping2007a; @wahlinMethodAnalysisGene2004].

The  recent introduction of lower cost and high throughput single cell sequencing technology has led to an explosion of research across many fields. As of January 2021, over 40 million cells have been sequenced across over 1,200 studies and the average size of each study past 2020 is over 100,000 cells [@svenssonCuratedDatabaseReveals2020]. The retina was used as the source tissue in one of the earliest works in the high throughput single cell transcriptomics field [@macoskoHighlyParallelGenomewide2015] and as of late 2020, over a dozen published studies containing over a million cells have used single cell technology to profile cell type specific gene expression patterns, cell fate trajectory, tissue and cell differentiation, and disease perturbation across multiple mammalian species [@buenaventuraIdentificationGenesEnriched2019; @clarkComprehensiveAnalysisRetinal2018; @collinCRXExpressionPluripotent2019; @collinDeconstructingRetinalOrganoids2019; @cowanCellTypesHuman2020; @diacouSix3Six6Are2018; @eldredThyroidHormoneSignaling2018; @hengHypoxiaToleranceNorrindeficient2019; @huDissectingTranscriptomeLandscape2019; @logiudiceSinglecellTranscriptionalLogic2019; @lukowskiSinglecellTranscriptomeAtlas2019; @luSingleCellAnalysisHuman2020;  @menonSinglecellTranscriptomicAtlas2019; @okorenMicroglialFunctionDistinct2019; @pengMolecularClassificationComparative2019; @phillipsNovelApproachSingle2018; @rheaumeSingleCellTranscriptome2018; @ruedaHippoPathwayBlocks2019; @shekharComprehensiveClassificationRetinal2016; @sridharSingleCellTranscriptomicComparison2020; @tranSingleCellProfilesRetinal2019; @voigtMolecularCharacterizationFoveal2019; @voigtSinglecellTranscriptomicsHuman2019; @yanCellAtlasHuman2020].

@@

```{r fig1, fig.width=12, fig.height=6, fig.cap=fig1_cap, echo=FALSE, message = FALSE, warning = FALSE}
fig1_cap <- fig_cap(name='fig1', caption =  'a. Cartoon schematic of retina with major cell types delinearted b. Simplified directed workflow of major steps in scEiaD creation from raw counts to gene counts, benchmarking optimal integration methods (SnakePOP) to produce batch corrected latent dimensions (LatentDims), then downstream analysis outputs like clustering, differential gene testing (DiffTesting), and 2D UMAP visualization. c. Counts of (study) accession, published papers, and batches for each scRNA technlogy, split by organism. d. Cell type counts extracted from published studies for the more common retina cell types, split by species. Count of study accessions for each species overlaid on bar plot.')

source('figs_and_tables/fig1.R')
retina_cartoon <- ggdraw() + draw_image("figs_and_tables/retina_cells_flat2.png", scale = 1)
plot_grid(
  plot_grid(retina_cartoon, b + coord_cartesian(clip = "off"), nrow = 1, labels = c('','b'), rel_widths = c(0.5,1)), 
  plot_grid(a, c, align = 'v',  axis = 'lr', nrow = 1, labels = c('','d')),
  nrow = 2, 
  rel_heights= c(1,1), 
  labels = c('a','c'))
```

@@

## Running `r methods %>% length()` tools in a Snakemake-based system

@@

```{r figBenchmark, fig.width=14, fig.height=8, fig.cap=figBenchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
figBenchmark_cap <- fig_cap(name='figBenchmark', caption =  'a. Example of a method (combat) which has a high level of batch blending, but poor separation of cell types (colored by cell type). b. no batch correction cleanly separates cell types but does not mix batches (colored by study). c. sumZScale (higher is better) for each method across a variety of data normalizations. All methods shown here use 2000 HVG, louvain clustering, and 8 latent dimensions d. Optimal parameters for scVI methods across 1000 to 10000 HVGs and 4 to 100 latent dimensions.')

@@

```{r figXGboost_umap, fig.width=18, fig.height=13, fig.cap=figXGboost_umap_cap, echo=FALSE, message = FALSE, warning = FALSE}
figXGboost_umap_cap <- fig_cap(name='figXGboost_umap', caption =  'a. Top genes that are differentially expressed across the major cell types of the retina. Genes are colored by which cell type they are differentially expressed in. The dot size is proportional to the percentage of those cells that have detectable levels of the gene. The color of the dot is the log2 scaled CPM expression. b. 2D UMAP projection of scEiaD, colored by cell type. The Tabula Muris data is colored all in gray. c. Facet plot that demonstrates how each major cell type of the retina is contained within a distinct space. d. Confusion matrix of cell type prediction performance of our xgboost labeller between predicted (y axis) and known (x axis) with data withheld from the machine learner. Most of the cell types are indeed labelled as their true type. e. Faceting of 2D UMAP by species and colored by cell type demonstrate how the major cell types of the retina share space with like cell types, despite being from mouse, human, and macaque.')

@@

input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 1
input[['pt_size_meta']] <- 1
#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['label_toggle']] <- 2
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- NULL
ctp <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)

input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 2
input[['pt_size_meta']] <- 2
input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- c('AC/HC Precursors','Amacrine Cells','Astrocytes','Bipolar Cells','Cones','Early RPCs','Endothelial','Horizontal Cells','Late RPCs','Microglia','Muller Glia','Neurogenic Cells','Pericytes','PR Precursors','Red Blood Cells','Retinal Ganglion Cells','Rod Bipolar Cells','Rods','RPCs','RPE','Schwann','Tabula Muris')
ctp2 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                    celltype_predict_labels,
                                    celltype_labels,
                                    tabulamuris_predict_labels,
                                    cluster_labels,
                                    cat_to_color_df
)

input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 1
input[['pt_size_meta']] <- 1
#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- NULL
org <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)

@@

After ML projection of cell type labels from the original `r meta_filter %>% filter(!is.na(CellType)) %>% nrow()` labels onto a total of `r meta_filter %>% filter(!is.na(CellType_predict) | !is.na(TabulaMurisCellType_predict)) %>% nrow()` cells we have substantially improved the number of studies per cell type. For example, we went from `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(2)` human studies with labelled Muller Glia to `r joinedSA %>% filter(CellType == 'Muller Glia') %>% pull(5)` after labeling (`r supTab_cap(name = 'supTab_SAcounts', display = 'cite')`). Overall we go from an average of `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(2) %>% mean()` studies per human cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(5) %>% mean()` and `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(4) %>% mean()` studies per mouse cell type to `r joinedSA %>% filter(CellType != 'Unlabelled') %>% pull(7) %>% mean()` after transferring the cell type labels.

As another check on the quality of the cell type assignments, we ran pairwise differential expression tests between the predicted cell types (see methods for further details). We show the five most consistently differentially expressed genes for each of the major retina cell types (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a). As a simple metric to identify known and unknown genes relating to the cell type specific expression we search PubMed for the number of publications with two searches per gene. We expect most of the genes identified to be known in the literature. The first search is the more precise "gene AND cell type" (e.g. "PDE6H AND Cones") and the second search is the more inclusive "gene AND retina" (e.g. "PDE6G AND Retina"). Of the 55 genes in (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`a), 40 had one or more citations in the gene - cell type search (table X) and 51 had one or more citation in the inclusive search. The 55 genes had a mean of `r top_markers %>% group_by(cluster)  %>% slice_max(mean_auc, n =5) %>% pull(c2) %>% mean() %>% round()` studies (with the inclusive gene by "retina" search). In contrast, 100 randomly chosen genes had a mean of `r pmid3 %>% map(function(x) sum(!is.na(x))) %>% unlist() %>% enframe() %>% pull(value) %>% mean() %>% round`.

@@

The photoreceptors (cones and rods) of the retina which are responsible for color and low-light vision, respectively, are near each other in the UMAP space. The major remaining retina cell types, by proportion in the mammalian eye are the muller glia, which are a glial cell type which help support the neurons of the retina. Next we have the neural cell types which transmit and help interpret the signals from the photoreceptors before they leave the retina via the optic stalk: the amacrine cells, retinal ganglia, horizontal, and bipolar cells. All of these cells are in well defined spaces in the UMAP. Finally we see even across species that the major cell types still group near each other (`r fig_cap(name = 'figXGboost_umap', display = 'cite')`d). The macaque retina cells are not present in the precursor/neurogenic center of the UMAP as expected because only fully developed tissues were sampled in these data.

@@

Many retina "atlases" have been published to date. We argue that we have created the first atlas that is generally useful because 1. our dataset/atlas is several times larger than any other published set, 2. our data is available via download in several forms at https://github.com/davemcg/scEiaD, and crucially 3. we provide a Google colab/Jupyter notebook which exactly lays out how to use scVI's to project (or query) outside data onto our scEiaD with minimum compute resources. Finally we also provide our xgboost model so outside groups can label their own datasets with our robust retina cell type labeling system.

# Supplemental Information
## Tables
```{r supTab1_CTcounts, fig.width=8, fig.height=14, out.width=300, echo = F, results='asis', warning = F, fig.cap = supTab1_CTcounts_cap, fig.pos = 'H'}
supTab1_CTcounts_cap <- supTab_cap(name='supTab1_CTcounts_cap', caption =  'Counts for cell type labels. Published are the author created labels from the published datasets. Transferred are the cell labels that were transferred by our xgboost-based machine learning model onto the entire scEiaD dataset.')
cat("<div id=\"tbl:supTab1_CTcounts\">")
source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
fontsize(ctTable, size = 8, part = 'all')
#Output Caption (* * for italics)
cat(paste0("\n*",supTab1_CTcounts_cap,"*\n"))
# Output </div>
cat("</div>")
```

```{r supTab_CTmislabels, fig.width=2, fig.cap=supTab_CTmislabels_cap, echo=FALSE, results='asis', message = FALSE, message = FALSE, warning = FALSE}

supTab_CTmislabels_cap <- supTab_cap(name='supTab_CTmislabels', caption =  'Counts of cell type labels with our xgboost machine learning system (PredCellType) and the published cell type labels (TrueCellType).')

cat("<div id=\"tbl:supTab_CTmislabels\">")
#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
fontsize(predictions %>%  
           mutate(max_pred_prob = rowMaxs(.[,-(31:ncol(predictions))] %>% as.matrix )) %>% 
           select(-cell_type_id) %>% 
           rename(PredCellType = CellType) %>% 
           inner_join(umapRef %>% select(Barcode, TrueCellType = CellType)) %>% 
           filter(!is.na(TrueCellType)) %>% 
           mutate(pred_correct  = ifelse(PredCellType == TrueCellType, 'Correct', 'incorrect')) %>% 
           group_by(PredCellType, TrueCellType) %>% 
           summarise(Count = n()) %>% 
           mutate(Ratio  = round(Count / sum(Count),2 )) %>% 
           filter(Ratio > 0.05) %>% 
           filter(!grepl('Doubl|Periocu', TrueCellType)) %>% 
           flextable() %>% 
           autofit(), 
         size = 8, 
         part = 'all')
#Output Caption (* * for italics)
cat(paste0("\n*",supTab_CTmislabels_cap,"*\n"))
# Output </div>
cat("</div>")
```

```{r supTab_SAcounts, fig.width=8, fig.height=14, out.width=300, fig.cap=supTab_SAcounts_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
cat("<div id=\"tbl:supTab_SAcounts\">")
supTab_SAcounts_cap <- supTab_cap(name='supTab_SAcounts_cap', caption =  'Counts for number of studies with cell types labels before and after cell type label transfer')

source('figs_and_tables/sup_fig__xgboost_PR.R')
fontsize(saTable, size = 8, part = 'all')
#Output Caption (* * for italics)
cat(paste0("\n*",supTab_SAcounts_cap,"*\n"))
# Output </div>
cat("</div>")
```

```{r supTab_AUCxgboost, fig.cap=supTab_AUCxgboost_cap, echo=FALSE, results='asis', message = FALSE, warning = FALSE}
supTab_AUCxgboost_cap <- supTab_cap(name='supTab_AUCxgboost', caption =  'Area under the precision recall curve (AUC) for each cell type, split by study')

cat("<div id=\"tbl:supTab_AUCxgboost\">")
#source('figs_and_tables/sup_fig2__celltype_xgboost_labelling.R')
fontsize(xgboost_pr_table_full %>% arrange(`Cell Type`, Study) %>% flextable() %>% width(width = 3), size = 8, part = 'all')
#Output Caption (* * for italics)
cat(paste0("\n*",supTab_AUCxgboost_cap,"*\n"))
# Output </div>
cat("</div>")
```

## Figures

```{r supFig_benchmark, fig.width=12, fig.height=8, out.width=300, fig.cap=supFig_benchmark_cap, echo=FALSE, message = FALSE, warning = FALSE}
supFig_benchmark_cap <- supFig_cap(name='supFig_benchmark', caption =  'Performance of the various batch correction tools across various benchmarking metrics. For the LISI and Silhouette plots in A, B higher (y-axis) means better batch mixing and further to the right (x-axis) means better cluster purity. For the ARI and NMI metrics (which reflects how well cluster matches with cell type) in C, D, higher means a better score.')

@@

```{r supFig_platformInt, fig.width=12, fig.height=10,  fig.cap=supFig_platformInt_cap, echo=FALSE, message = FALSE, warning = FALSE}
supFig_platformInt_cap <- supFig_cap(name='supFig_platformInt', caption =  'Distribution of cell types across the 2D UMAP confirms that the cell types are being properly placed despite the wide variety of single sequencing platforms present.')
input <- list()
input[['meta_column']] <- 'CellType_predict'
input[['pt_size_back']] <- 5
input[['pt_size_meta']] <- 5
input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
#input[['meta_filter_cat']] <- 'CellType_predict'
#input[['meta_filter_on']] <- 'Bipolar Cells'
ctp3 <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                    celltype_predict_labels,
                                    celltype_labels,
                                    tabulamuris_predict_labels,
                                    cluster_labels,
                                    cat_to_color_df
)

@@

```{r projection_fig, fig.width=12, fig.height=10,  fig.cap=projection_fig_cap, echo=FALSE, message = FALSE, warning = FALSE}
projection_fig_cap <- supFig_cap(name='projection_fig', caption =  'a. RPE distribution colored by study demonstrates how the RPE are grouped in a similar location. iPSC-based RPE we provided are located within the tissue based RPE cells. b. Screen shot of Google colab notebook that demonstrates how to integrate external data into the scEiaD resource. We see in the screenshort how the dataset contains many of the retinal cell types.')
input[['pt_size_back']] <- 0
input[['pt_size_meta']] <- 6
input[['meta_column']] <- 'study_accession'
#input[['gene_and_meta_scatter_tech']] <- 'Droplet'
input[['meta_column_transform']] <- 'None'
input[['meta_filter_cat']] <- 'CellType_predict'
input[['meta_filter_on']] <- c('RPE')
rpe <- make_meta_scatter_umap_plot(input, mf, meta_filter,
                                   celltype_predict_labels,
                                   celltype_labels,
                                   tabulamuris_predict_labels,
                                   cluster_labels,
                                   cat_to_color_df
)
#

@@

```{r supFig1, fig.width=4, fig.height=6, out.width=300, fig.cap=supFig1_cap, echo=FALSE, message = FALSE, warning = FALSE}
supFig1_cap <- supFig_cap(name='supFig1_cap', caption =  'The left bar delineates the number of cells for each organism - technology combination. The right bar specifies the number of each cells in each post QC category. In silico doublets were identified with scrublet and DoubletDetector.')

```{r supFig_PR, fig.width=12, fig.height=10,  fig.cap=supFig_PR_cap, echo=FALSE, message = FALSE, warning = FALSE}
supFig_PR_cap <- supFig_cap(name='supFig_PR', caption =  'Precision recall curves for our xgboost cell type predictor model across each cell type predicted')

pr_split_curves
```

# Works Cited
